WASP-4 is Accelerating Towards the Earth

### 2020-01-28 paper/ms.tex
update paper

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period is decreasing.
  Potential causes of the period change include tidal orbital decay,
  orbital precession, and light-travel time effects.
  In this work we present new radial velocity measurements of WASP-4,
  acquired with Keck-HIRES.
  The new data show that the system is accelerating towards the Earth
  at $\dot{\gamma} = X.XX \pm Y.YY\ {\rm m}\,{\rm s}^{-1}\,{\rm
  yr}^{-1}$.
  The corresponding period decrease explains WASP-4's
  changing orbital period as a light-travel time effect.
  Combining the radial velocities with new speckle imaging limits, we
  find that system's acceleration is likely caused by a
  10-200$\,M_{\rm Jup}$ companion with semi-major axis between 10-100$\,$AU.
  Based on statistics from \citet{knutson_friends_2014}, about 1 in 6
  hot Jupiters are expected to show period decreases commensurate with
  WASP-4 due to acceleration by outer companions.
  Continued long-term radial velocity monitoring of hot Jupiters is
  therefore essential to distinguish tidal orbital decay from
  line-of-sight accelerations.
\end{abstract}

@@

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.  We discovered the timing variations using
data from TESS and a decade of ground-based monitoring
\citep[][hereafter \citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.
Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  With additional
data, they found a lower best-fit decay rate of $\dot{P} = -XX.X \pm
X.X$ milliseconds per year.
A separate study by \citet{baluev_2019} reported
additional archival transit light curves of WASP-4b, most notably from
the TRAPPIST archive, and a select few other observers.
% \citeauthor{baluev_2019} analyzed their newly obtained
% photometry, along with archival light curves that we omitted from our
% analysis due to systematic uncertainties in the absolute time system.
% \citeauthor{baluev_2019} found that when they used all the available
% TTV data, the need for a quadratic ephemeris was present.  However,
% they pointed out that if they used lower-precision subsets of the
% available timing data, the necessity for the quadratic term decreased.
% This is not particularly surprising.
% \citeauthor{baluev_2019} also pointed out that the
% precise transit times reported by \citet{huitson_gemini_2017} were
% quite important in the time-series.  Overall,
% \citeauthor{baluev_2019} did not find the claim of a
% period decrease convincing.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the radial velocities
were sparse and showed a weak ($<2\sigma$) linear trend.  Our new
measurements reveal a line-of-sight acceleration of $\dot{\gamma} =
-XX {\rm m\,s^{-1}\,yr^{-1}}$.  This translates to an expected period
decrease simply from the light travel time effect of ZZ milliseconds
per year --- about commensurate with what is observed.
In the following, we avoid the term ``R{\o}mer delay'' as this phrase
typically signifies arrival time delay due to observatory motion. We
are discussing an analog of the Doppler effect for a source with
constant line-of-sight acceleration.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed times minus the calculated
  times assuming a constant orbital period.  More opaque points correspond
  to more precise data.  The quadratic ephemeris and its $1\sigma$
  statistical uncertainties are shown in blue.  Neither model fully
  explains the transit residuals (see~\ref{sec:transit_analysis}).
  The binned TESS point (yellow star) is the weighted average of 18
  TESS transits and is for display purposes only.  Its uncertainty is
  slightly larger than the marker size.  The models were fitted to all
  of the individual transit times.
  \label{fig:times}
	}
\end{figure}

We considered two possible models for the observed transit times.  The
first model assumes a constant orbital period on a circular orbit:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,\\
\end{align}
where $E$ is the epoch number and $t_0$ is a reference epoch.  The
second model assumes the period changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
\end{align}
The free parameters are the reference epoch $t_0$, the period at the
reference epoch, and the period derivative, ${\rm d}P/{\rm d}t = (1/P)
{\rm d}P/{\rm d}E$.  We defined the epoch numbers such that $E=0$ is
near the weighted average of the observed times.  This helps to reduce
the covariance between $t_0$ and $P$.  A third possible model that we
did not consider for reasons that will become apparent is a
precessing, eccentric orbit \citep[{\it
e.g.},][]{gimenez_revision_1995,patra_2017}.

@@

%FIXME: perhaps include the WASP-4b phase-fold here too??
\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity observations of WASP-4.} The best-fit Keplerian orbit
    of WASP-4b has been subtracted.  The linear trend inferred from the
    RV data is shown with a black line, with $1\sigma$ errors in gray.
    The trend that would be needed to produce the period decrease seen
    in transits ($\dot{P}= - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$) is
    indicated with the purple dotted line.  The four new RV
    measurements presented in this work increased the significance of
    the slope from $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

WASP-4 was found to be accelerating towards our line-of-sight at high
confidence,
\begin{equation}
  \dot{v}_{\rm r} =
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
For comparison, before our recent observing run, 
$\dot{v}_{\rm r}$ was thought to be about five times smaller, and was
only marginally significant
\citep{knutson_friends_2014,bouma_wasp4b_2019}.

@@

%TODO: describe 3b better
%TODO: also, put 3b in as the bottom right inset to the contrast curve
\begin{figure}[!t]
	\gridline{
		\fig{f3a.png}{0.33\textwidth}{}
	}
	\vspace{-1cm}
	\gridline{
		\fig{f3b.pdf} {0.5\textwidth}{}
	}
	\vspace{-0.7cm}
    \caption{
      {\it Top:\,} 
      Speckle image reconstructed from 1000 40 millisecond frames in
      an 832 nm bandpass, and acquired on September 28, 2019.  The
      image scale is $2.46''\times2.46''$.
      {\it Bottom:\,} 
      Contrast curve derived from point-source injection-recovery
      experiments.
    }
    \label{fig:zorro}
\end{figure}

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Masses and semi-major axes of companions that meet
    requirements of both the radial velocity and the speckle imaging.}
    The likelihood inferred from radial velocities is shown in
    grayscale, and the region excluded from the imaging is shown with
    a cross-hatch pattern.
	\label{fig:mass_sma}
  \vspace{-0.3cm}
	}
\end{figure}

@@

We began by defining a $75\times50$ grid in true planetary mass and
semimajor axis, with even logarithmic spacing from 1 to 900$\,$$M_{\rm
Jup}$ and 3 to 500$\,$${\rm AU}$.  We then considered the possibility
that an additional companion in any particular cell could explain the
observed linear trend.  In each cell, we simulated 500 hypothetical
companions.

@@

For each simulated companion, we then drew a sample from the converged
chains of our initial model of WASP-4b, plus its linear trend. We
subtracted the planet's orbital solution, leaving the linear trend.
Given $(a_{\rm c}, M_{\rm c}, e_{\rm c}$ for each simulated outer
companion, and the fixed instrument offsets and jitters from the MCMC
chains, we then performed a maximum likelihood fit for the time and
argument of periastron of the outer simulated companion.  We converted
the resulting $75\times50\times500$ cube of best-fit log-likelihood
values to probabilities, and averaged over the samples in each grid
cell to derive a probability distribution in mass and semi-major axis.

@@

>>>>

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period is decreasing.
  Potential causes of the period change include tidal orbital decay,
  orbital precession, and light-travel time effects.
  In this work we present new radial velocity measurements of WASP-4,
  acquired with Keck-HIRES.
  The new data show that the system is accelerating towards the Earth
  at $\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm m}\,{\rm
  s}^{-1}\,{\rm yr}^{-1}$.
  The corresponding period decrease explains WASP-4's
  changing orbital period as a light-travel time effect.
  Combining the radial velocities with new speckle imaging limits, we
  find that system's acceleration is likely caused by a
  10-200$\,M_{\rm Jup}$ companion with semi-major axis between 10-100$\,$AU.
  Based on statistics from \citet{knutson_friends_2014}, about 1 in 6
  hot Jupiters are expected to show period decreases commensurate with
  WASP-4 due to acceleration by outer companions.
  Continued long-term radial velocity monitoring of hot Jupiters is
  therefore essential to distinguish tidal orbital decay from
  line-of-sight accelerations.
\end{abstract}

@@

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.  We discovered the timing variations using
data from TESS and a decade of ground-based monitoring
\citep[][hereafter \citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.
Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  With additional
data, they found a lower best-fit decay rate of $\dot{P} = -XX.X \pm
X.X$ milliseconds per year.
A separate study by \citet{baluev_2019} reported
additional archival transit light curves of WASP-4b, most notably from
the TRAPPIST archive, and a select few other observers.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the radial velocities
were sparse and showed a weak ($<2\sigma$) linear trend.  Our new
measurements reveal a line-of-sight acceleration of $\dot{\gamma} =
-0.0422^{+0.0028}_{-0.0027}\ {\rm m\,s^{-1}\,yr^{-1}}$.  This
translates to an expected period
decrease simply from the light travel time effect of ZZ milliseconds
per year --- about commensurate with what is observed.
In the following, we avoid the term ``R{\o}mer delay'' as this phrase
typically signifies arrival time delay due to observatory motion. We
are discussing an analog of the Doppler effect for a source with
constant line-of-sight acceleration.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed transit times minus the calculated
  times assuming a constant orbital period.  More opaque points correspond
  to more precise data.  The quadratic ephemeris and its $\pm1\sigma$
  statistical uncertainties are shown in blue.  Neither model fully
  explains the scatter in transit residuals, perhaps due to
  underestimated uncertainties (see~\ref{sec:transit_analysis}).
  The binned TESS point (yellow star) is the weighted average of 18
  TESS transits and is binned for display purposes only.  Its uncertainty is
  slightly larger than the marker size.  The models were fitted to all
  of the individual transit times.
  \label{fig:times}
	}
\end{figure}

We considered two possible models for the observed transit times.  The
first model assumes a constant orbital period on a circular orbit:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the epoch number and $t_0$ is a reference epoch.  The
second model assumes the period changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
\end{align}
The free parameters are the reference epoch $t_0$, the period at the
reference epoch, and the period derivative, ${\rm d}P/{\rm d}t = (1/P)
{\rm d}P/{\rm d}E$.  We defined the epoch numbers such that $E=0$ is
near the weighted average of the observed times.  This helps to reduce
the covariance between $t_0$ and $P$.  A third possible model that we
did not consider for reasons that will become apparent is a
precessing, eccentric orbit \citep[{\it
e.g.},][]{gimenez_revision_1995,patra_2017}.

@@

%FIXME: perhaps include the WASP-4b phase-fold here too??
\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity observations of WASP-4.} The best-fit Keplerian orbit
    of WASP-4b has been subtracted.  The linear trend inferred from the
    RV data is shown with a black line, with $1\sigma$ errors in gray.
    The trend that would be needed to produce the period decrease seen
    in transits ($\dot{P}= - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$) is
    indicated with the purple dotted line.  The four new RV
    measurements presented in this work increased the significance of
    the radial velocity trend from $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

WASP-4 was found to be accelerating towards our line-of-sight at high
confidence,
\begin{equation}
  \dot{v}_{\rm r} = \dot{\gamma} = 
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
For comparison, before our recent observing run, 
$\dot{v}_{\rm r}$ was thought to be about five times smaller, and was
only marginally significant
\citep{knutson_friends_2014,bouma_wasp4b_2019}.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f3.pdf}
	\end{center}
	\vspace{-0.7cm}
    \caption{
      {\bf Zorro contrast curve derived from point-source
      injection-recovery experiments.}
      Sources below the curve would have been detected.
      The inset shows the speckle image reconstructed from 1000 40
      millisecond frames in an 832 nm bandpass, and acquired on
      September 28, 2019.  The image scale is $2.46''\times2.46''$.
    }
    \label{fig:zorro}
\end{figure}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.78\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Masses and semi-major axes of companions that meet
    requirements of both the radial velocity and the speckle imaging.}
    The likelihood inferred from radial velocities is shown in
    grayscale, and the region excluded from the imaging is shown with
    a cross-hatch pattern.
	\label{fig:mass_sma}
  \vspace{-0.3cm}
	}
\end{figure*}

@@

We began by defining a $128\times128$ grid in true planetary mass and
semimajor axis, with even logarithmic spacing from 1 to 900$\,$$M_{\rm
Jup}$ and 3 to 500$\,$${\rm AU}$.  We then considered the possibility
that an additional companion in any particular cell could explain the
observed linear trend.  In each cell, we simulated 512 hypothetical
companions.

@@

For each simulated companion, we then drew a sample from the converged
chains of our initial model of WASP-4b, plus its linear trend. We
subtracted the planet's orbital solution, leaving the linear trend.
Given $(a_{\rm c}, M_{\rm c}, e_{\rm c}$ for each simulated outer
companion, and the fixed instrument offsets and jitters from the MCMC
chains, we then performed a maximum likelihood fit for the time and
argument of periastron of the outer simulated companion.  We converted
the resulting $128\times128\times512$ cube of best-fit log-likelihood
values to probabilities, and averaged over the samples in each grid
cell to derive a probability distribution in mass and semi-major axis.

@@

% TODO: add a paragraph noting that while the skepticism
% is fine "I was right".

% \citeauthor{baluev_2019} analyzed their newly obtained
% photometry, along with archival light curves that we omitted from our
% analysis due to systematic uncertainties in the absolute time system.
% \citeauthor{baluev_2019} found that when they used all the available
% TTV data, the need for a quadratic ephemeris was present.  However,
% they pointed out that if they used lower-precision subsets of the
% available timing data, the necessity for the quadratic term decreased.
% This is not particularly surprising.
% \citeauthor{baluev_2019} also pointed out that the
% precise transit times reported by \citet{huitson_gemini_2017} were
% quite important in the time-series.  Overall,
% \citeauthor{baluev_2019} did not find the claim of a
% period decrease convincing.
